Location and Scale-Invariant Power Transformations for Transforming Data to Normality

### 2024-10-18 manuscript_latex/manuscript.tex
Add citations in Bibtex format.

%% The hyperref package allows for typesetting hyperlinks.
%% \usepackage{hyperref}

@@

Many statistical and some machine learning methods assume normality of
the underlying data, e.g.~analysis of variance and linear discriminant
analysis. However, numerical features in datasets may strongly deviate
from normal distributions, e.g.~by being skewed. Power transformations
aim to stabilise variance and improve normality of such features
(Bartlett 1947; John W. Tukey 1957). The two most commonly used
transformations are that of Box and Cox (1964) and Yeo and Johnson
(2000). The Box-Cox transformation of a feature value \(x_i\) of feature
\(\mathbf{X}=\left\{x_1, x_2, \ldots, x_n \right\}\) under the
transformation parameter \(\lambda\) is defined as:

@@

The \(\lambda\)-parameter is typically optimised using maximum
likelihood estimation under the assumption that the transformed feature
is normally distributed. As noted by Raymaekers and Rousseeuw, this
approach is sensitive to outliers, and robust versions of Box-Cox and
Yeo-Johnson transformations were devised (Raymaekers and Rousseeuw
2024).

@@

Statistical tests for normality, such as the Shapiro-Wilk test (Shapiro
and Wilk 1965), could be automatically applied to transformed features.
However, given sufficiently large sample sizes such tests can detect
trivial deviations from normality, and may lead to rejection of
sufficiently good power transformations.

@@

{\centering \includegraphics{figure_1}

@@

Real-world data may contain outliers, to which maximum likelihood
estimation can be sensitive. Their presence may lead to poor
transformations to normality, as shown in Figure
\ref{fig:decreased-normality}. As indicated by Raymaekers and Rousseeuw
(2024), the general aim of power transformations should be to transform
non-outlier data to normality, i.e.~achieve \emph{central normality}. To
achieve this, they devised an iterative procedure to find a robust
estimate of the transformation parameter \(\lambda\). Briefly, this
process requires identifying outliers in the data and weighting such
instances during the optimisation process. Raymaekers and Rousseeuw
(2024) achieve this through weighted maximum likelihood estimation.
However, because this procedure iteratively estimates and updates
\(\lambda\), it can not be used here to simultaneously estimate
\(\lambda\), \(x_0\) and \(s\) for location- and scale-invariant power
transformations. Nonetheless, as a procedure, weighted MLE can be used
for estimating the transformation, shift and scale parameters.

Here, weighted maximum likelihood estimation is based on equations
\ref{eqn:box-cox-invariant-log-likelihood} and
\ref{eqn:yeo-johnson-invariant-log-likelihood}. Compared to Raymaekers
and Rousseeuw (2024), these log-likelihood functions includes additional
terms to accommodate estimation of \(x_0\) and \(s\). The weighted
location- and scale-invariant Box-Cox log-likelihood function is:

@@

\begin{itemize}
\item
  A tapered cosine function (J. W. Tukey 1967), with \(\delta_1 \geq 0\)
  and \(\delta_2 \geq \delta_1\) as threshold parameters:
\end{itemize}

@@

\begin{itemize}
\item
  The weighting function uses empirical probabilities of the
  distribution of the original feature \(\mathbf{X}\). After sorting
  \(\mathbf{X}\) in ascending order, probabilities are determined as
  \(p_i = \frac{i - 1/3}{n + 1/3}\), with \(i = 1, 2, \ldots n\), with
  \(n\) the number of instances of feature \(\mathbf{X}\). Then
  \(\dot{x}_i = p^{*}_i=2 \left( p_i - 0.5\right)\), so that argument is
  zero-centered.
\item
  The weighting function uses the z-score of the transformed feature
  \(\phi^{\lambda, x_0, s} (\mathbf{X})\). After Raymaekers and
  Rousseeuw (2024),
  \(z_i = \frac{\phi^{\lambda, x_0, s}(x_i) - \mu_M}{\sigma_M}\). Here,
  \(\mu_M\) and \(\sigma_M\) are robust Huber M-estimates of location
  and scale of the transformed feature
  \(\phi^{\lambda, x_0, s} (\mathbf{X})\) (Huber 1981). Then
  \(\dot{x}_i = z_i\).
\item
  After sorting \(\mathbf{X}\) in ascending order, the weighting
  function uses the residual error between the z-score of the
  transformed feature \(\phi^{\lambda, x_0, s} (\mathbf{X})\) and the
  theoretical z-score from a standard normal distribution:
  \(r_i =\left| \left( \phi^{\lambda, x_0, s}(x_i) - \mu_M)\right) / \sigma_M - F^{-1}_{\mathcal{N}}(p_i) \right|\),
  with \(\mu_M\), \(\sigma_M\) and \(p_i\) as defined above. Then
  \(\dot{x}_i = r_i\).
\end{itemize}

@@

Modifications intended to make power transformations invariant to
location and scale of a feature and methods to improve their robustness
against outliers need to be assessed using data drawn from a range of
different distributions. Since the power transformations are intended
for use with unimodal distributions, the generalised normal distribution
(Subbotin 1923; Nadarajah 2005) is a suitable option for simulating
realistic feature distributions. This distribution has the following
probability density function \(f_{\beta}\) for a value
\(x \in \mathbb{R}\):

@@

Realistic feature distributions may be skewed. Gijbels et al.~describe a
recipe for introducing skewness into the otherwise symmetric generalised
normal distribution (Gijbels, Karim, and Verhasselt 2019), leading to
the following probability density function:

@@

The quantile function for the asymmetric generalised normal distribution
\(F_{\alpha}^{-1}\) thus incorporates the quantile function
\(F_{\beta}^{-1}\) of the symmetric generalised normal distribution.
\(F_{\beta}^{-1}\) was derived by Griffin to be (Griffin 2018):

@@

Power transformations aim to transform features to a normal
distribution. However, this may not always be successful or possible.
Deviations from normality can be detected by normality tests, such as
the Shapiro-Wilk test (Shapiro and Wilk 1965). In practice, normality
tests may be too stringent with large sample sizes, outliers, or both.
Here we develop an empirical test for central normality. The null
hypothesis \(H_0\) is that the distribution is centrally normal. The
alternative hypothesis \(H_1\) is that the distribution is not centrally
normal.

Let central normality be defined as the normality of central portion
\(\kappa\) of the data,
i.e.~\(\mathbf{X}_{\text{central}} = \left\{x_i \in \mathbf{X} \, | \,  \frac{1-\kappa}{2} \leq  p_i \leq \frac{1 + \kappa}{2}\right\}\),
with \(p_i\) probabilities of the empirical distribution, as previously.
We then compute the residual errors between the z-scores of the
transformed feature \(\phi^{\lambda, x_0, s} (\mathbf{X})\) and the
expected z-scores from a standard normal distribution:
\(r_i =\left| \left( \phi^{\lambda, x_0, s}(x_i) - \mu_M)\right) / \sigma_M - F^{-1}_{\mathcal{N}}(p_i) \right|\),
with \(\mu_M\) and \(\sigma_M\) robust Huber M-estimates of location and
scale of the transformed feature \(\phi^{\lambda, x_0, s} (\mathbf{X})\)
(Huber 1981). The set of residual errors for the central portion of the
data is then
\(\mathbf{R}_{\text{central}} = \left\{ r_i \in \left\{ r_1, r_2, \ldots, r_n\right\} \, | \,  \frac{1-\kappa}{2} \leq  p_i \leq \frac{1 + \kappa}{2}\right\}\).

@@

We used simulated data to assess invariance to location and scale of the
proposed power transformations, weighting for robust transformations,
and to develop the empirical central normality test. The \(\lambda\)
parameter for conventional power transformations (Eqn.
\ref{eqn:box-cox-original} and \ref{eqn:yeo-johnson-original}), as well
as \(\lambda\), \(x_0\) and \(s\) parameters for location- and
scale-invariant power transformations (Eqn. \ref{eqn:box-cox-invariant}
and \ref{eqn:yeo-johnson-invariant}) were estimated using the BOBYQA
algorithm for derivative-free bound constraint optimisation (Powell
2009) through maximum likelihood estimation. The required algorithms
were implemented in the \texttt{power.transform} R software package
(Zwanenburg and Löck 2024b) (version 1.0.0). Of note, the
\texttt{power.transform} package shifts feature values into the positive
domain if negative or zero values are present for Box-Cox power
transformations.

@@

Outlier values were then drawn to randomly replace a fraction of the
values of \(\mathbf{X}_i\). This was repeated \(m_{\text{out}} = 10\)
times, with outlier fractions regularly spaced in \([0.00, 0.10]\). Thus
up to 10 percent of the values could be replaced by outliers. Outlier
values were set according to John W. Tukey (1977), as follows. Let
\(x^{*} \sim U\left(-2, 2\right)\). Then the corresponding outlier value
was:

@@

Minimisation was conducted using the BOBYQA algorithm for
derivative-free bound constraint optimisation (Powell 2009). The
resulting weighting function parameters for weighted MLE are shown in
Tables \ref{tab:optimal-weighting-parameters-box-cox} and
\ref{tab:optimal-weighting-parameters-yeo-johnson} for robust location-
and scale-invariant Box-Cox and Yeo-Johnson transformations,
respectively.

@@

A common feature in health-related datasets is age. Here we use data on
228 patients with lung cancer that was collected and published by
Loprinzi et al. (Loprinzi et al. 1994). The age in the cohort was
\(62.4 \pm 9.1\) (mean ± standard deviation) years. Applying
conventional and invariant Yeo-Johnson transformations to patient age
yielded the following results, see Figure
\ref{fig:experimental-results-invariance}: no transformation (sum of
residuals with normal distribution \(\sum r_i = 16.5\)); conventional
transformation (\(\lambda = 2.0\), \(\sum r_i = 11.5\),
\(\mu_{YJ} = 1.8 \cdot 10^3\), \(\sigma_{YJ} = 0.5 \cdot 10^3\));
Raymaekers and Rousseeuw's robust adaptation (\(\lambda = 2.0\),
\(\sum r_i = 11.5\), \(\mu_{YJ} = 1.8 \cdot 10^3\),
\(\sigma_{YJ} = 0.5 \cdot 10^3\)); location- and scale-invariant
transformation (\(\lambda = 0.9\), \(\sum r_i = 8.8\),
\(\mu_{YJ} = 1.2\), \(\sigma_{YJ} = 1.1\)); and robust location- and
scale-invariant transformation (\(\lambda = 0.8\), \(\sum r_i = 10.6\),
\(\mu_{YJ} = 1.2\), \(\sigma_{YJ} = 1.0\)).

@@

Gorman, Williams and Fraser recorded body mass (in grams) of 342
penguins of three different species (Gorman, Williams, and Fraser 2014).
The body mass was \((4.2 \pm 0.8) \cdot 10^3\) (mean ± standard
deviation) grams, and not centrally normal (\(p = 0.03\)). Applying
conventional and invariant Yeo-Johnson transformations to body mass
yielded the following results, see Figure
\ref{fig:experimental-results-invariance}: no transformation (residual
sum \(\sum r_i = 48.0\)); conventional transformation
(\(\lambda = -0.5\), \(\sum r_i = 32.2\), \(\mu_{YJ} = 2.1\),
\(\sigma_{YJ} = 4 \cdot 10^{-3}\)); Raymaekers and Rousseeuw's robust
adaptation (\(\lambda = -0.5\), \(\sum r_i = 32.2\), \(\mu_{YJ} = 2.1\),
\(\sigma_{YJ} = 4 \cdot 10^{-3}\)); location- and scale-invariant
transformation (\(\lambda = 0.5\), \(\sum r_i = 26.8\),
\(\mu_{YJ} = 0.9\), \(\sigma_{YJ} = 0.9\)); and robust location- and
scale-invariant transformation (\(\lambda = 0.4\), \(\sum r_i = 23.1\),
\(\mu_{YJ} = 0.8\), \(\sigma_{YJ} = 0.8\)).

@@

Geospatial datasets usually contain coordinates. The Ames housing
dataset contains data on 2930 properties that were sold between 2006 and
2010 (De Cock 2011), including their geospatial coordinates. The
latitude was \(42.03 \pm 0.02)\) (mean ± standard deviation). Applying
conventional and invariant Yeo-Johnson transformations to latitude
yielded the following results, see Figure
\ref{fig:experimental-results-invariance}: no transformation (residual
sum \(\sum r_i = 328\)); conventional transformation
(\(\lambda = 62.1\), \(\sum r_i = 319\),
\(\mu_{YJ} = 4.8 \cdot 10^{99}\), \(\sigma_{YJ} = 0.1 \cdot 10^{99}\));
Raymaekers and Rousseeuw's robust adaptation (\(\lambda = 95.4\),
\(\sum r_i = 319\), \(\mu_{YJ} = 6.4 \cdot 10^{153}\),
\(\sigma_{YJ} = 0.3 \cdot 10^{153}\)); location- and scale-invariant
transformation (\(\lambda = 1.5\), \(\sum r_i = 326\),
\(\mu_{YJ} = -1.2\), \(\sigma_{YJ} = 0.8\)); and robust location- and
scale-invariant transformation (\(\lambda = 1.4\), \(\sum r_i = 311\),
\(\mu_{YJ} = -1.3\), \(\sigma_{YJ} = 0.9\)).

@@

The Top Gear dataset contains data on 297 vehicles that appeared on the
BBC television show \emph{Top Gear} (Alfons 2021). Within this dataset,
the fuel consumption feature contains outliers due to highly
fuel-efficient vehicles. Applying conventional and invariant Yeo-Johnson
transformations to the fuel consumption feature yielded the following
results, see Figure \ref{fig:experimental-results-outlier-robustness}:
no transformation (residual sum \(\sum r_i = 54\), \(p=0.76\));
conventional transformation (\(\lambda = -0.1\), \(\sum r_i = 55\),
\(\mu_{YJ} = 3.0\), \(\sigma_{YJ} = 0.3\), \(p=0.01\)); Raymaekers and
Rousseeuw's robust adaptation (\(\lambda = 0.8\), \(\sum r_i = 48\),
\(\mu_{YJ} = 29\), \(\sigma_{YJ} = 15\), \(p=0.55\)); location- and
scale-invariant transformation (\(\lambda = -1.3\), \(\sum r_i = 44\),
\(\mu_{YJ} = 0.5\), \(\sigma_{YJ} = 0.1\), \(p=0.03\)); and robust
location- and scale-invariant transformation (\(\lambda = -0.9\),
\(\sum r_i = 50\), \(\mu_{YJ} = 2.0\), \(\sigma_{YJ} = 2.3\),
\(p=0.58\)).

@@

The ischemic stroke dataset contains historic data from 126 patients
with risk at ischemic stroke (Kuhn and Johnson 2019). These patients
underwent Computed Tomography Angiography to characterize the carotid
artery blockages. Angiography imaging was then assessed, and various
characteristics related to the blood vessels and the disease are
measured. The maximum arterial wall thickness feature contains several
instances with outlier values. Applying conventional and invariant
Yeo-Johnson transformations to this feature yielded the following
results, see Figure \ref{fig:experimental-results-outlier-robustness}:
no transformation (residual sum \(\sum r_i = 110\), \(p=0.56\));
conventional transformation (\(\lambda = -0.7\), \(\sum r_i = 30\),
\(\mu_{YJ} = 1.0\), \(\sigma_{YJ} = 0.1\), \(p=0.01\)); Raymaekers and
Rousseeuw's robust adaptation (\(\lambda = 1.1\), \(\sum r_i = 136\),
\(\mu_{YJ} = 7.2\), \(\sigma_{YJ} = 14\), \(p=0.61\)); location- and
scale-invariant transformation (\(\lambda = 0.2\), \(\sum r_i = 12\),
\(\mu_{YJ} = -11.8\), \(\sigma_{YJ} = 6.9\), \(p=0.13\)); and robust
location- and scale-invariant transformation (\(\lambda = -0.6\),
\(\sum r_i = 27\), \(\mu_{YJ} = 0.7\), \(\sigma_{YJ} = 0.1\),
\(p=0.10\)).

@@

We used 285 datasets from the Penn Machine Learning Benchmarks
collection (Romano et al. 2022). In this collection, 122 datasets
correspond to regression tasks and 163 datasets to classification tasks.
Using the familiar auto-machine learning library (Zwanenburg and Löck
2024a) (version 1.5.0), each dataset was used to train a model for each
of 16 process configurations. Each process configuration specifies the
learner (generalised linear model or random forest), transformation
method (none, conventional Yeo-Johnson, robust invariant Yeo-Johnson,
robust invariant Yeo-Johnson with empirical central normality test
(rejecting transformations with \(p \leq 0.01\)), and normalisation
method (none, \(z\)-standardisation), yielding 16 distinct
configurations. Before each experiment, each dataset was randomly split
into a training (70\%) and holdout test (30\%) set five times. Thus, a
total of 22800 models were created. Each model was then evaluated using
the holdout test set using one of two metrics, i.e.~the root relative
squared error (RRSE) for regression tasks and the area under the
receiver operating characteristic curve (AUC) for classification tasks.

@@

To estimate the marginal effects of process parameters, including
transformation method, we first fit a regression random forest (ranger
package (Wright and Ziegler 2017) version 0.16.0): 2000 trees, minimum
node size 2, other hyperparameters default) with process parameters and
task difficulty as predictors and normalised rank as response variable.
The estimated marginal effects are shown in Figure
\ref{fig:marginal-effect-plot}. On the scale of normalised ranks
(\([0.0, 1.0]\)), the overall estimated marginal effect of using a
random forest instead of generalised linear model was \(0.272\). The
overall marginal effect of using z-standardisation to normalise features
was \(0.008\). Transformation methods had the following marginal
effects: \(0.003\) for using conventional Yeo-Johnson transformation
instead of no transformation; \(-0.007\) for using robust invariant
Yeo-Johnson transformation instead of no transformation; \(-0.009\) for
using robust invariant Yeo-Johnson transformation with empirical central
normality test instead of no transformation; \(-0.010\) for using robust
invariant Yeo-Johnson transformation instead of conventional Yeo-Johnson
transformation; and \(-0.012\) for using using robust invariant
Yeo-Johnson transformation with empirical central normality test instead
of conventional Yeo-Johnson transformation.

@@

Similarly to Raymaekers and Rousseeuw (2024), sample weights \(w_i\) are
introduced to facilitate robust power transformations. The weighted
log-likelihood of the location- and scale-invariant Box-Cox
transformation is:

>>>>

@@

Many statistical and some machine learning methods assume normality of
the underlying data, e.g.~analysis of variance and linear discriminant
analysis. However, numerical features in datasets may strongly deviate
from normal distributions, e.g.~by being skewed. Power transformations
aim to stabilise variance and improve normality of such features
\citep{Bartlett1947-rx, Tukey1957-rt}. The two most commonly used
transformations are that of \citet{Box1964-mz} and \citet{Yeo2000-vw}.
The Box-Cox transformation of a feature value \(x_i\) of feature
\(\mathbf{X}=\left\{x_1, x_2, \ldots, x_n \right\}\) under the
transformation parameter \(\lambda\) is defined as:

@@

The \(\lambda\)-parameter is typically optimised using maximum
likelihood estimation under the assumption that the transformed feature
is normally distributed. As noted by Raymaekers and Rousseeuw, this
approach is sensitive to outliers, and robust versions of Box-Cox and
Yeo-Johnson transformations were devised \citep{Raymaekers2024-zf}.

@@

Statistical tests for normality, such as the Shapiro-Wilk test
\citep{Shapiro1965-zd}, could be automatically applied to transformed features.
However, given sufficiently large sample sizes such tests can detect
trivial deviations from normality, and may lead to rejection of
sufficiently good power transformations.

@@

{\centering \includegraphics[width=1\linewidth]{figure_1}

@@

Real-world data may contain outliers, to which maximum likelihood
estimation can be sensitive. Their presence may lead to poor
transformations to normality, as shown in Figure
\ref{fig:decreased-normality}. As indicated by \citet{Raymaekers2024-zf},
the general aim of power transformations should be to transform
non-outlier data to normality, i.e.~achieve \emph{central normality}. To
achieve this, they devised an iterative procedure to find a robust
estimate of the transformation parameter \(\lambda\). Briefly, this
process requires identifying outliers in the data and weighting such
instances during the optimisation process. \citet{Raymaekers2024-zf}
achieve this through weighted maximum likelihood estimation.
However, because this procedure iteratively estimates and updates
\(\lambda\), it can not be used here to simultaneously estimate
\(\lambda\), \(x_0\) and \(s\) for location- and scale-invariant power
transformations. Nonetheless, as a procedure, weighted MLE can be used
for estimating the transformation, shift and scale parameters.

Here, weighted maximum likelihood estimation is based on equations
\ref{eqn:box-cox-invariant-log-likelihood} and
\ref{eqn:yeo-johnson-invariant-log-likelihood}. Compared to \citet{Raymaekers2024-zf},
these log-likelihood functions includes additional
terms to accommodate estimation of \(x_0\) and \(s\). The weighted
location- and scale-invariant Box-Cox log-likelihood function is:

@@

\begin{itemize}
\item
  A tapered cosine function \citep{Tukey1967-eb}, with \(\delta_1 \geq 0\)
  and \(\delta_2 \geq \delta_1\) as threshold parameters:
\end{itemize}

@@

\begin{itemize}
\item
  The weighting function uses empirical probabilities of the
  distribution of the original feature \(\mathbf{X}\). After sorting
  \(\mathbf{X}\) in ascending order, probabilities are determined as
  \(p_i = \frac{i - 1/3}{n + 1/3}\), with \(i = 1, 2, \ldots n\), with
  \(n\) the number of instances of feature \(\mathbf{X}\). Then
  \(\dot{x}_i = p^{*}_i=2 \left( p_i - 0.5\right)\), so that argument is
  zero-centered.
\item
  The weighting function uses the z-score of the transformed feature
  \(\phi^{\lambda, x_0, s} (\mathbf{X})\). After \citep{Raymaekers2024-zf},
  \(z_i = \frac{\phi^{\lambda, x_0, s}(x_i) - \mu_M}{\sigma_M}\). Here,
  \(\mu_M\) and \(\sigma_M\) are robust Huber M-estimates of location
  and scale of the transformed feature
  \(\phi^{\lambda, x_0, s} (\mathbf{X})\) \citep{Huber1981-su}. Then
  \(\dot{x}_i = z_i\).
\item
  After sorting \(\mathbf{X}\) in ascending order, the weighting
  function uses the residual error between the z-score of the
  transformed feature \(\phi^{\lambda, x_0, s} (\mathbf{X})\) and the
  theoretical z-score from a standard normal distribution:
  \(r_i =\left| \left( \phi^{\lambda, x_0, s}(x_i) - \mu_M)\right) / \sigma_M - F^{-1}_{\mathcal{N}}(p_i) \right|\),
  with \(\mu_M\), \(\sigma_M\) and \(p_i\) as defined above. Then
  \(\dot{x}_i = r_i\).
\end{itemize}

@@

Modifications intended to make power transformations invariant to
location and scale of a feature and methods to improve their robustness
against outliers need to be assessed using data drawn from a range of
different distributions. Since the power transformations are intended
for use with unimodal distributions, the generalised normal distribution
\citep{Subbotin1923-qk, Nadarajah2005-xe} is a suitable option for simulating
realistic feature distributions. This distribution has the following
probability density function \(f_{\beta}\) for a value
\(x \in \mathbb{R}\):

@@

Realistic feature distributions may be skewed. Gijbels et al.~describe a
recipe for introducing skewness into the otherwise symmetric generalised
normal distribution \citep{Gijbels2019-te}, leading to
the following probability density function:

@@

The quantile function for the asymmetric generalised normal distribution
\(F_{\alpha}^{-1}\) thus incorporates the quantile function
\(F_{\beta}^{-1}\) of the symmetric generalised normal distribution.
\(F_{\beta}^{-1}\) was derived by Griffin to be \citep{Griffin2018-bf}:

@@

Power transformations aim to transform features to a normal
distribution. However, this may not always be successful or possible.
Deviations from normality can be detected by normality tests, such as
the Shapiro-Wilk test \citep{Shapiro1965-zd}. In practice, normality
tests may be too stringent with large sample sizes, outliers, or both.
Here we develop an empirical test for central normality. The null
hypothesis \(H_0\) is that the distribution is centrally normal. The
alternative hypothesis \(H_1\) is that the distribution is not centrally
normal.

Let central normality be defined as the normality of central portion
\(\kappa\) of the data,
i.e.~\(\mathbf{X}_{\text{central}} = \left\{x_i \in \mathbf{X} \, | \,  \frac{1-\kappa}{2} \leq  p_i \leq \frac{1 + \kappa}{2}\right\}\),
with \(p_i\) probabilities of the empirical distribution, as previously.
We then compute the residual errors between the z-scores of the
transformed feature \(\phi^{\lambda, x_0, s} (\mathbf{X})\) and the
expected z-scores from a standard normal distribution:
\(r_i =\left| \left( \phi^{\lambda, x_0, s}(x_i) - \mu_M)\right) / \sigma_M - F^{-1}_{\mathcal{N}}(p_i) \right|\),
with \(\mu_M\) and \(\sigma_M\) robust Huber M-estimates of location and
scale of the transformed feature \(\phi^{\lambda, x_0, s} (\mathbf{X})\)
\citep{Huber1981-su}. The set of residual errors for the central portion of the
data is then
\(\mathbf{R}_{\text{central}} = \left\{ r_i \in \left\{ r_1, r_2, \ldots, r_n\right\} \, | \,  \frac{1-\kappa}{2} \leq  p_i \leq \frac{1 + \kappa}{2}\right\}\).

@@

We used simulated data to assess invariance to location and scale of the
proposed power transformations, weighting for robust transformations,
and to develop the empirical central normality test. The \(\lambda\)
parameter for conventional power transformations (Eqn.
\ref{eqn:box-cox-original} and \ref{eqn:yeo-johnson-original}), as well
as \(\lambda\), \(x_0\) and \(s\) parameters for location- and
scale-invariant power transformations (Eqn. \ref{eqn:box-cox-invariant}
and \ref{eqn:yeo-johnson-invariant}) were estimated using the BOBYQA
algorithm for derivative-free bound constraint optimisation \citep{Powell2009-zb}
through maximum likelihood estimation. The required algorithms
were implemented in the \texttt{power.transform} R software package
(version 1.0.0) \citep{Zwanenburg2024-kq}. Of note, the
\texttt{power.transform} package shifts feature values into the positive
domain if negative or zero values are present for Box-Cox power
transformations.

@@

Outlier values were then drawn to randomly replace a fraction of the
values of \(\mathbf{X}_i\). This was repeated \(m_{\text{out}} = 10\)
times, with outlier fractions regularly spaced in \([0.00, 0.10]\). Thus
up to 10 percent of the values could be replaced by outliers. Outlier
values were set according to \citet{Tukey1977-xm}, as follows. Let
\(x^{*} \sim U\left(-2, 2\right)\). Then the corresponding outlier value
was:

@@

Minimisation was conducted using the BOBYQA algorithm for
derivative-free bound constraint optimisation \citep{Powell2009-zb}. The
resulting weighting function parameters for weighted MLE are shown in
Tables \ref{tab:optimal-weighting-parameters-box-cox} and
\ref{tab:optimal-weighting-parameters-yeo-johnson} for robust location-
and scale-invariant Box-Cox and Yeo-Johnson transformations,
respectively.

@@

A common feature in health-related datasets is age. Here we use data on
228 patients with lung cancer that was collected and published by
Loprinzi et al. \citep{Loprinzi1994-cd}. The age in the cohort was
\(62.4 \pm 9.1\) (mean ± standard deviation) years. Applying
conventional and invariant Yeo-Johnson transformations to patient age
yielded the following results, see Figure
\ref{fig:experimental-results-invariance}: no transformation (sum of
residuals with normal distribution \(\sum r_i = 16.5\)); conventional
transformation (\(\lambda = 2.0\), \(\sum r_i = 11.5\),
\(\mu_{YJ} = 1.8 \cdot 10^3\), \(\sigma_{YJ} = 0.5 \cdot 10^3\));
Raymaekers and Rousseeuw's robust adaptation (\(\lambda = 2.0\),
\(\sum r_i = 11.5\), \(\mu_{YJ} = 1.8 \cdot 10^3\),
\(\sigma_{YJ} = 0.5 \cdot 10^3\)); location- and scale-invariant
transformation (\(\lambda = 0.9\), \(\sum r_i = 8.8\),
\(\mu_{YJ} = 1.2\), \(\sigma_{YJ} = 1.1\)); and robust location- and
scale-invariant transformation (\(\lambda = 0.8\), \(\sum r_i = 10.6\),
\(\mu_{YJ} = 1.2\), \(\sigma_{YJ} = 1.0\)).

@@

Gorman, Williams and Fraser recorded body mass (in grams) of 342
penguins of three different species \citep{Gorman2014-eo}.
The body mass was \((4.2 \pm 0.8) \cdot 10^3\) (mean ± standard
deviation) grams, and not centrally normal (\(p = 0.03\)). Applying
conventional and invariant Yeo-Johnson transformations to body mass
yielded the following results, see Figure
\ref{fig:experimental-results-invariance}: no transformation (residual
sum \(\sum r_i = 48.0\)); conventional transformation
(\(\lambda = -0.5\), \(\sum r_i = 32.2\), \(\mu_{YJ} = 2.1\),
\(\sigma_{YJ} = 4 \cdot 10^{-3}\)); Raymaekers and Rousseeuw's robust
adaptation (\(\lambda = -0.5\), \(\sum r_i = 32.2\), \(\mu_{YJ} = 2.1\),
\(\sigma_{YJ} = 4 \cdot 10^{-3}\)); location- and scale-invariant
transformation (\(\lambda = 0.5\), \(\sum r_i = 26.8\),
\(\mu_{YJ} = 0.9\), \(\sigma_{YJ} = 0.9\)); and robust location- and
scale-invariant transformation (\(\lambda = 0.4\), \(\sum r_i = 23.1\),
\(\mu_{YJ} = 0.8\), \(\sigma_{YJ} = 0.8\)).

@@

Geospatial datasets usually contain coordinates. The Ames housing
dataset contains data on 2930 properties that were sold between 2006 and
2010 \citep{De-Cock2011-jf} including their geospatial coordinates. The
latitude was \(42.03 \pm 0.02)\) (mean ± standard deviation). Applying
conventional and invariant Yeo-Johnson transformations to latitude
yielded the following results, see Figure
\ref{fig:experimental-results-invariance}: no transformation (residual
sum \(\sum r_i = 328\)); conventional transformation
(\(\lambda = 62.1\), \(\sum r_i = 319\),
\(\mu_{YJ} = 4.8 \cdot 10^{99}\), \(\sigma_{YJ} = 0.1 \cdot 10^{99}\));
Raymaekers and Rousseeuw's robust adaptation (\(\lambda = 95.4\),
\(\sum r_i = 319\), \(\mu_{YJ} = 6.4 \cdot 10^{153}\),
\(\sigma_{YJ} = 0.3 \cdot 10^{153}\)); location- and scale-invariant
transformation (\(\lambda = 1.5\), \(\sum r_i = 326\),
\(\mu_{YJ} = -1.2\), \(\sigma_{YJ} = 0.8\)); and robust location- and
scale-invariant transformation (\(\lambda = 1.4\), \(\sum r_i = 311\),
\(\mu_{YJ} = -1.3\), \(\sigma_{YJ} = 0.9\)).

@@

The Top Gear dataset contains data on 297 vehicles that appeared on the
BBC television show \emph{Top Gear} \citep{Alfons2021-kc}. Within this dataset,
the fuel consumption feature contains outliers due to highly
fuel-efficient vehicles. Applying conventional and invariant Yeo-Johnson
transformations to the fuel consumption feature yielded the following
results, see Figure \ref{fig:experimental-results-outlier-robustness}:
no transformation (residual sum \(\sum r_i = 54\), \(p=0.76\));
conventional transformation (\(\lambda = -0.1\), \(\sum r_i = 55\),
\(\mu_{YJ} = 3.0\), \(\sigma_{YJ} = 0.3\), \(p=0.01\)); Raymaekers and
Rousseeuw's robust adaptation (\(\lambda = 0.8\), \(\sum r_i = 48\),
\(\mu_{YJ} = 29\), \(\sigma_{YJ} = 15\), \(p=0.55\)); location- and
scale-invariant transformation (\(\lambda = -1.3\), \(\sum r_i = 44\),
\(\mu_{YJ} = 0.5\), \(\sigma_{YJ} = 0.1\), \(p=0.03\)); and robust
location- and scale-invariant transformation (\(\lambda = -0.9\),
\(\sum r_i = 50\), \(\mu_{YJ} = 2.0\), \(\sigma_{YJ} = 2.3\),
\(p=0.58\)).

@@

The ischemic stroke dataset contains historic data from 126 patients
with risk at ischemic stroke \citep{Kuhn2019-kt}. These patients
underwent Computed Tomography Angiography to characterize the carotid
artery blockages. Angiography imaging was then assessed, and various
characteristics related to the blood vessels and the disease are
measured. The maximum arterial wall thickness feature contains several
instances with outlier values. Applying conventional and invariant
Yeo-Johnson transformations to this feature yielded the following
results, see Figure \ref{fig:experimental-results-outlier-robustness}:
no transformation (residual sum \(\sum r_i = 110\), \(p=0.56\));
conventional transformation (\(\lambda = -0.7\), \(\sum r_i = 30\),
\(\mu_{YJ} = 1.0\), \(\sigma_{YJ} = 0.1\), \(p=0.01\)); Raymaekers and
Rousseeuw's robust adaptation (\(\lambda = 1.1\), \(\sum r_i = 136\),
\(\mu_{YJ} = 7.2\), \(\sigma_{YJ} = 14\), \(p=0.61\)); location- and
scale-invariant transformation (\(\lambda = 0.2\), \(\sum r_i = 12\),
\(\mu_{YJ} = -11.8\), \(\sigma_{YJ} = 6.9\), \(p=0.13\)); and robust
location- and scale-invariant transformation (\(\lambda = -0.6\),
\(\sum r_i = 27\), \(\mu_{YJ} = 0.7\), \(\sigma_{YJ} = 0.1\),
\(p=0.10\)).

@@

We used 285 datasets from the Penn Machine Learning Benchmarks
collection \citep{Romano2022-gq}. In this collection, 122 datasets
correspond to regression tasks and 163 datasets to classification tasks.
Using the familiar auto-machine learning library (version 1.5.0) \citep{Zwanenburg2021-so},
each dataset was used to train a model for each
of 16 process configurations. Each process configuration specifies the
learner (generalised linear model or random forest), transformation
method (none, conventional Yeo-Johnson, robust invariant Yeo-Johnson,
robust invariant Yeo-Johnson with empirical central normality test
(rejecting transformations with \(p \leq 0.01\)), and normalisation
method (none, \(z\)-standardisation), yielding 16 distinct
configurations. Before each experiment, each dataset was randomly split
into a training (70\%) and holdout test (30\%) set five times. Thus, a
total of 22800 models were created. Each model was then evaluated using
the holdout test set using one of two metrics, i.e.~the root relative
squared error (RRSE) for regression tasks and the area under the
receiver operating characteristic curve (AUC) for classification tasks.

@@

To estimate the marginal effects of process parameters, including
transformation method, we first fit a regression random forest (ranger
package version 0.16.0, \citep{Wright2017-rf}): 2000 trees, minimum
node size 2, other hyperparameters default) with process parameters and
task difficulty as predictors and normalised rank as response variable.
The estimated marginal effects are shown in Figure
\ref{fig:marginal-effect-plot}. On the scale of normalised ranks
(\([0.0, 1.0]\)), the overall estimated marginal effect of using a
random forest instead of generalised linear model was \(0.272\). The
overall marginal effect of using z-standardisation to normalise features
was \(0.008\). Transformation methods had the following marginal
effects: \(0.003\) for using conventional Yeo-Johnson transformation
instead of no transformation; \(-0.007\) for using robust invariant
Yeo-Johnson transformation instead of no transformation; \(-0.009\) for
using robust invariant Yeo-Johnson transformation with empirical central
normality test instead of no transformation; \(-0.010\) for using robust
invariant Yeo-Johnson transformation instead of conventional Yeo-Johnson
transformation; and \(-0.012\) for using using robust invariant
Yeo-Johnson transformation with empirical central normality test instead
of conventional Yeo-Johnson transformation.

@@

Similarly to \citet{Raymaekers2024-zf}, sample weights \(w_i\) are
introduced to facilitate robust power transformations. The weighted
log-likelihood of the location- and scale-invariant Box-Cox
transformation is:
